---
title: "Challenge 7: Incorporating Multiple Inputs"
subtitle: "Functions + Fish"
author: "Lisette Abundez"
format: html
editor: source
embed-resources: true
code-tools: true
code-fold: true
execute: 
  message: false
  warning: false
  echo: true
---

```{r}
#| label: setup

library(here)
library(tidyverse)

fish <- read_csv(here("data", "BlackfootFish.csv"))
```

## Condition Index

A frequently used measurement for fish health is a condition index [(Wikipedia article)](https://en.wikipedia.org/wiki/Condition_index_in_fish). The following
simple equation can be used to calculate the approximate condition index of a
fish:

$$\text{condition index} = \frac{weight}{length^3} \times 100$$

**1. There are specific units required for the calculation of a condition index -- length must be in centimeters and weight must be in grams. The weight data for the Blackfoot River fish were collected in grams; the length data were collected in millimeters. Transform the length data to the correct units.**

```{r}
#| label: transform-data-to-have-correct-units

fish <- fish |>
  mutate(length = length / 10)

```

**2. Collecting data of this sort can be very messy! Write a function that will replace unlikely `length` and `weight` measurements with `NA`. Your function should accept at least three inputs:**

+ **a vector of measurements,**
+ **the minimum reasonable value,**
+ **the maximum reasonable value.**

**If a value falls outside these bounds, you should replace it with an `NA`.**

::: callout-tip
If you are struggling with the structure of your function, I would suggest reading the [Mutating Function](https://r4ds.hadley.nz/functions.html#mutate-functions) from R4DS.
:::

```{r}
#| label: function-to-change-impossible-measurements-to-NA

library(dplyr)

unlikely <- function(x, min, max) {
  dplyr::case_when(
    x < min ~ NA,  
    x > max ~ NA,  
    .default = x          
  )
}
```

**3. After consulting the [Montana Record Table](https://fwp.mt.gov/fish/anglingData/records) for the four species of trout included in these data, I have conjectured that it is unlikely to have measurements for fish below 5 cm and above 80 cm in length or below 10 g and above 4,000 g in weight. Use your function to modify the `length` and `weight` columns of the `fish` dataset based on my cutoffs.**

```{r}
#| label: modify-impossible-values-of-length-weight
fish <- fish |>
  mutate(
    length = unlikely(length, min = 5, max = 80),
    weight = unlikely(weight, min = 10, max = 4000)
  )
```

**3. Write a function to calculate the condition index of a fish, given inputs of weight and length.**

::: callout-warning
Consider whether you will write a function that works with vectors (accepting
vectors as inputs) or a function that works with data frames (accepting a 
dataset and variable names as inputs)!
:::

```{r}
#| label: condition-function

condition_index <- function(weight, length) {
  CI <- (weight / (length^3)) * 100
  return(CI)
}
```

**4. Make a thoughtful visualization of how fish conditions have varied over the duration of this study.**

::: callout-warning
### To be thoughtful...

... you need to use at least a third variable!
:::

```{r}
#| label: condition-indices-over-time

library(ggplot2)

fish <- fish |>
  mutate(condition_index = condition_index(length, weight)) 

ggplot(fish, aes(x = year, y = condition_index, fill = species)) +
  geom_bar(stat = "identity", position = "stack") +  
  labs(x = "Year",
       y = NULL,
       subtitle = "Condition Index",
       title = "Condition Index by Species Over Time",
       fill = "Species") +
  theme_minimal()

```
